---
title: "Predicting Missing Values"
description: |
  Analising forest inventory data and predicting NAs through least squares regression model
author:
  - name: Gabriel de Freitas Pereira
    url: {}
date: 04-18-2021
output:
  distill::distill_article:
    self_contained: false
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(knitr)
library(kableExtra)
library(dplyr)

```


\  
The sistematic collection of data was carried out between 2017 and 2018 at Três Lagoas, 
which is a city from Mato Grosso do Sul state. The area has an extent of 1226 acres of *Eucalyptus grandis* W. Hill ex Maiden x *Eucalyptus urophylla* S. T. Blake 
clonal culture with mean age of 2.88 years. 


```{r echo = FALSE, caption}
fustes <- read.csv2("fustes.csv")
fustes_6 <- fustes %>% 
  select(area, espacamento, matgen,
         idade, dap, ht) 

names(fustes_6) <- c("area(ha)", "spacing", "genetic material",
                     "age", "DBH", "heigth")  

```
    
# Exploring the data
\  
\ 

\  
Here we can take a look on the first 6 rows from the data frame which has 3313 obs. of 6 variables:

```{r, echo = FALSE}
knitr::kable(head(fustes_6), booktabs= TRUE) %>% 
  kable_styling(position = 'left', full_width = F)

```
\  
\   

  
Checking whether we have got NAs or not:

```{r, echo = TRUE, background= "blue"}
sum(is.na(fustes_6$DBH))

```
\  
\  

\  
As we can see there are 134 NAs from DBH values that probably represent dead trees, that can be replanted in the future for 
next rotation. But just to be sure about it, let's take a look on the age (or height)! Because if
those NAs values also could be trees too much young for DBH.

```{r, echo = TRUE}
min(fustes_6$age)

```
\ 
\  


\  
The minimum age turns out the idea of NAs are representing dead trees. 
Therefore if we would like to get a second rotation forestry would be good to understand the behavior of trees under similar conditions to do future economic analysis. 
So what would we expect from these trees if they were alive?
\  

\  
We can make an approximation through least squares regression model based on our data and variables to find out their diameter at breast height!
And for that we need to pre-process the data... 
\  
\  

# Data pre-processing 
\  
\  

Checking whether we have got outliers on DBH or not:

```{r, echo = FALSE}
boxplot(fustes_6$DBH, main = "Diameter at Breast Height")

```

\  
As we can see, there are outliers on the data that we want to 
remove because it's not good for least squares regression model.
The result bellow show us the lower extreme, first quartile, median, third quartile and
the upper extreme respectively. 

```{r, echo = TRUE}
boxplot.stats(fustes_6$DBH)

```
\  
\  

  
Automating the process to get the extreme values:  

```{r, echo = TRUE}
upper_whisker <- boxplot.stats(fustes_6$DBH)$stats[5]
lower_whisker <- boxplot.stats(fustes_6$DBH)$stats[1]

```
\  
\  

  
Of course after the upper extreme we've got the outliers and before the first whisker as well. And both we are going to remove this way:

```{r,echo = TRUE}

outlier_filter_upper <- fustes_6$DBH < upper_whisker 
outlier_filter_lower <- fustes_6$DBH > lower_whisker 

#filtering outliers: 
outlier_filter <- fustes_6[outlier_filter_lower & outlier_filter_upper,]

```
\ 
\  

  
Now we need to check the class from the variables and if necessary make some changes 

```{r, echo = TRUE}
str(fustes_6)

```
\ 
\  

  
Categorical casting for the variables that we will use to predict

```{r, echo = TRUE}
fustes_6$`genetic material` <- as.factor(fustes_6$`genetic material`)
```
\ 
\  
  
# Building the model
\  
\  
  
Our formula has the following format: y = b0 + b1X

```{r, echo = TRUE}
dbh_equation <- "DBH ~ `genetic material` + age"

```
\  
\  
  
  
Least squares regression:

```{r, echo = TRUE}
dbh_model <- lm(
  formula = dbh_equation, 
  data = fustes_6[outlier_filter_lower & outlier_filter_upper,]
)

```
\  
\ 

  
Now we are going to isolate the NA values (that we want to predict) and 
all the columns that we are using to get DBH values:

```{r, echo = TRUE}
dbh_rows <- fustes_6[is.na(fustes_6$DBH), c("genetic material","age")]
```
\  
\ 
 
  
The last step is to predict and replace those missing values

```{r, echo = TRUE}
dbh_predictions <- predict(dbh_model, newdata = dbh_rows)
dbh_predictions

```

Filling the data with predicted values:

```{r, echo = TRUE}
fustes_6[is.na(fustes_6$DBH), "DBH"] <- dbh_predictions

```


